---
title: 'Statistical Inference: Peer Graded Assigment - Part 2'
author: "Wayne Heller"
date: "April 22, 2017"
output:
  word_document: default
  html_document: default
  pdf_document:
    fig_caption: yes
    fig_height: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff=65), tidy=TRUE, fig.pos = 'H', fig.show = 'asis')
```

# OVERVIEW
*Part 2 Basic Inferential Data Analysis:*  This part of the analysis uses the ToothGrowth data in the R datasets library and compares tooth growth across dosage and delivery methods.  The specific instructions of this assignemnt are to "Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose"

## Description of the data
The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice coded as OJ or ascorbic acid a form of vitamin C and coded as VC.

``` {r echo=FALSE}
        library(ggplot2)
        library(datasets)
```

# EXPLORATORY ANALYSIS
``` {r echo=FALSE,  fig.width=5, fig.height=3, fig.align='center', fig.show='hold'}
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
createFig1 <- function() {
        g <- ggplot(data=ToothGrowth, aes(y=len, x=dose, fill=dose)) + geom_boxplot()
        g <- g + labs(title = "Average Tooth Growth Increases with Dose", x="Dose", y="Length", subtitle="Figure 1")
        print(g)
}
createFig1()

createFig2 <- function() {
        g <- ggplot(data=ToothGrowth, aes(y=len, x=supp, fill=supp)) + geom_boxplot()
        g <- g + facet_grid(. ~ dose)
        g <- g + labs(title = "For Lower Doses, Average Growth is Higher for OJ", x="Dose", y="Length", subtitle="Figure 2")
        print(g)
}
createFig2()

```

## Summary of Exploratory Analysis
1. The data indicate there maybe a strong relationship between an increase in tooth length and dose. Figure 1 on Left
2. The data indicate there maybe a difference in effectiveness between Orange Juice and  Vitamin C supplements. Figure 2 on Right

# HYPOTHESIS TESTING

### Hypothesis Test 1: Over all dosage levels, with 95% confidence is Orange Juice a more effective delivery method than Vitamin C supplements
**Assumptions:** Tooth growth is normally distributed  

**Approach:** Use Permutation testing to find percentage of results that produce a more extreme difference in averages than was measured  

**Ho:** `There is no significant difference in tooth growth between Orange Juice and Vitamin C`

**Ha:** `Average tooth growth is greater for Orange Juice` 

``` {r echo=TRUE}
y <- ToothGrowth$len
group <- ToothGrowth$supp
testStat <- function(l,g) mean(l[g=="OJ"])- mean(l[g=="VC"]) # test statistic difference of means between groups
observedStat <- testStat(y, group)                           # test statistic for real observations
permutations <- sapply(1:10000, function(i) testStat(y, sample(group))) # test statistics for permutations
observedStat
mean(permutations > observedStat)  # percentage of permutations that have a more extreme result than was observed
```

```{r echo=FALSE,  fig.cap="Figure 3: Permutations Analysis", fig.height=3}
createFig3 <- function() {
        dat <- data.frame(x=permutations)
        g <- ggplot(dat, aes(x = x, fill="red")) 
        g <- g + geom_histogram(alpha = .20, binwidth=1, color = "black") 
        g <- g + geom_vline(xintercept=observedStat, color='red') + theme(legend.position="none")
        g <- g + labs(title="Distribution of the Difference in Mean  OJ vs. VC", x="Test Statistic for Permutations")
        print(g)
}
createFig3()

```

##### Conclusion for Hypothesis 1
Based on the very low probability of finding a more extreme example, we should **REJECT the null hypothesis Ho:** `There is no significant difference in tooth growth between Orange Juice and Vitamin C`

### Hypothesis Test 2: At low dose (0.5), with 95% confidence is Orange Juice a more effective delivery method than Vitamin C supplements
**Assumptions:** Tooth growth is normally distributed  

**Ho:** `There is no difference in average growth`  

**Ha:** `Average tooth growth is greater for Orange Juice`

``` {r }
lenOJ <- ToothGrowth$len[ToothGrowth$dose==0.5 & ToothGrowth$supp=="OJ"]
lenVC <- ToothGrowth$len[ToothGrowth$dose==0.5 & ToothGrowth$supp=="VC"]
t.test(lenOJ, lenVC, alternative = "greater", paired = FALSE, var.equal = TRUE)$p.value
```

##### Conclusion for Hypothesis 2
Based on the Student's T Test, we should **REJECT the null Hypothesis Ho:** `There is no difference in average growth` 


### 3. At high dose (2.0), with 95% confidence is Orange Juice a more effective delivery method than Vitamin C supplements
**Assumptions:** Tooth growth is normally distributed  

**Ho:** `There is no difference in average growth`  

**Ha:** `Average tooth growth is greater for Orange Juice`

``` {r }
lenOJ <- ToothGrowth$len[ToothGrowth$dose==2.0 & ToothGrowth$supp=="OJ"]
lenVC <- ToothGrowth$len[ToothGrowth$dose==2.0 & ToothGrowth$supp=="VC"]
t.test(lenOJ, lenVC, alternative = "greater", paired = FALSE, var.equal = TRUE)$p.value
```
 
##### Conclusion for Hypothesis 3
Based on the Student's T Test, we should **ACCEPT the null Hypothesis Ho:** `There is no difference in average growth` 

# APPENDIX
### Code for Figure 1
``` {r echo=TRUE}
    createFig1
```

### Code for Figure 2
``` {r echo=TRUE}
    createFig2
```

### Code for Figure 3
``` {r echo=TRUE}
    createFig3
```