---
title: 'Statistical Inference: Peer Graded Assigment'
author: "Wayne Heller"
date: "April 22, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
Part 1 of this analysis compares a simulated sample mean and variance for an exponential distribution to the theoretical values, and to the means of 40 samples.

Part 2 is...

### Simulations
#### Exponential distribution with 1,000 simulated observations
```{r cache=TRUE}
        seed = 12345
        nosim <- 1000                 # number of simulated values
        lambda <-0.2
        x.dist <- rexp(nosim, lambda) # simulated exponential distribution
        dat <- data.frame(x = x.dist) # create a dataframe from the values for plotting
        x.mean = mean(dat$x)          # mean of the simulated distribution
        x.var = var(dat$x)            # variance of the simulated distribution
```

#### Distribution of Means from 1,000 simulations of 40 observations each
```{r cache=TRUE}
        nsize <- 40                   # number of observations in 1 simulation
        x.dist <- rexp(nosim * nsize, lambda) # simulated exponential distribution
        x.avg = apply(matrix(x.dist, nosim), 1, mean)
        dat.avg <- data.frame(x = x.avg)
        x.avg.mean = mean(dat.avg$x)  # simulated mean
        x.avg.var = var(dat.avg$x)    # simulated variance
        x.avg.mean.theoretical = 1/lambda # theoretical mean
        x.avg.var.theoretical = (1/lambda^2)/nsize
```

### Sample Mean and Variance vs. Theoretical
```{r echo=FALSE}
        library(ggplot2)
        g <- ggplot(dat, aes(x = x, fill="red")) 
        g <- g + geom_histogram(alpha = .20, binwidth=1, colour = "black") 
        g <- g + geom_vline(xintercept=x.mean) 
        g <- g + annotate(geom = "text", x=10, y=100, label=paste0("simulated mean=", as.character(round(x.mean,digits=2))), size=4, angle=0, vjust=0, hjust=0)
        
        g <- g + annotate(geom = "text", x=10, y=90, label=paste0("theoretical mean=", as.character(round(1/lambda,digits=2))), size=4, angle=0, vjust=0, hjust=0)
        
        g <- g + annotate(geom = "text", x=10, y=130, label=paste0("simulated variance=", as.character(round(x.var, digits=2))), size=4, angle=0, vjust=0, hjust=0)
        
        g <- g + annotate(geom = "text", x=10, y=120, label=paste0("theoretical variance=", as.character(round(1/lambda^2, digits=2))), size=4, angle=0, vjust=0, hjust=0)
        g <- g + theme(legend.position="none")
        g <- g + labs(title = paste("Simulated Exponential Distribution with lambda=", as.character(lambda), "and", as.character(nosim), "Observations"))
        print(g)
        
print(paste("The Sample Mean of", as.character(round(x.mean,digits=2)), "approximates the theoretical mean 1/lambda of", as.character(round(1/lambda, digits=2)) ))

print(paste("And the Sample Variance of", as.character(round(x.var,digits=2)), "approximates the theoretical variance 1/lambda^2 of", as.character(round(1/lambda^2, digits=2)) ))
```


### Distribution of Sample Means
``` {r echo=FALSE}
g <- ggplot(dat.avg, aes(x = x, fill='red')) 
g <- g + geom_histogram(alpha = .25, binwidth=.2, colour = "black", aes(y=..density..)) 
g <- g + geom_vline(xintercept=x.avg.mean, color="green", size = 1.0) + theme(legend.position="none")
g <- g + stat_function(fun=dnorm,args=list(mean=x.avg.mean, sd=sqrt(x.avg.var)),color = "green", size = 1.0)
g <- g + geom_vline(xintercept=x.avg.mean.theoretical, color="red", size=.5) + theme(legend.position="none")
g <- g + stat_function(fun=dnorm,args=list(mean=x.avg.mean.theoretical, sd=sqrt(x.avg.var.theoretical)),color = "red", size = 1.0)
g <- g + labs(title = "Density Distribution of 1,000 Means of 40  Observations", x= "Means of 40 Observations from a Simulated Exponential Distribution")
print(g)  

statistics <- c("Mean", "Variance")
theoretical <- c(x.avg.mean.theoretical, x.avg.var.theoretical)
simulated <- c(x.avg.mean, x.avg.var)
df <- data.frame(statistics, theoretical, simulated)
names(df) <- c("Statistic", "Theoretical Value", "Simulated Value")
library(knitr)
kable(df)
```

### Summary

## Part 2

### Exploratory Plots
``` {r echo=FALSE}
library(datasets)
ToothGrowth$dose <- as.factor(ToothGrowth$dose)

g <- ggplot(data=ToothGrowth, aes(y=len, x=dose, fill=dose)) + geom_boxplot()
g <- g + labs(title = "Average Tooth Growth Increases with Dose", x="Dose", y="Length")
print(g)

g <- ggplot(data=ToothGrowth, aes(y=len, x=supp, fill=supp)) + geom_boxplot()
g <- g + facet_grid(. ~ dose)
g <- g + labs(title = "For Lower Doses, Average Growth is Higher for OJ", x="Dose", y="Length")
print(g)

```

### Hypotheses to Test
#### At low dose (0.5), with 95% confidence is Orange Juice a more effective delivery mechanism than Vitamin C
Assumptions: Tooth growth is normally distributed  
Ho: There is no difference in average growth  
Ha: Average tooth growth is greater for Orange Juice

``` {r }
lenOJ <- ToothGrowth$len[ToothGrowth$dose==0.5 & ToothGrowth$supp=="OJ"]
lenVC <- ToothGrowth$len[ToothGrowth$dose==0.5 & ToothGrowth$supp=="VC"]
t.test(lenOJ, lenVC, alternative = "greater", paired = FALSE, var.equal = TRUE)$p.value
```