---
title: 'Statistical Inference: Peer Graded Assigment - Part 1'
author: "Wayne Heller"
date: "April 22, 2017"
output:
  word_document: default
  html_document: default
  pdf_document:
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff=65), tidy=TRUE, fig.pos = 'H', fig.show = 'asis')
```

## Overview
*Part 1 Central Limit Theorem Simulation:* This part of the analysis compares the sample mean and variance for a large number of means of a simulated exponential distribution.  The expected result is that the distribution of means will be Normal with a mean approximately equal to the theoretical mean of the exponential distribution and a variance approximately equal to the theoretical variance of the exponential distribution divided by the number of samples in each group. 

``` {r echo=FALSE}
        library(ggplot2)
        library(knitr)
```

## Simulations
Create an Exponential distribution with 1,000 simulated observations
```{r cache=TRUE }
        
        seed = 12345
        nosim <- 1000                 # number of simulated values
        lambda <-0.2                  # lambda of 0.2 was provided 
        x.dist <- rexp(nosim, lambda) # simulated exponential distribution
        dat <- data.frame(x = x.dist) # create a dataframe from the values for plotting
        x.mean = mean(dat$x)          # mean of the simulated distribution
        x.var = var(dat$x)            # variance of the simulated distribution
```

Next, create a distribution of Means from 1,000 simulations of 40 observations each
```{r cache=TRUE}
        nsize <- 40                                   # number of observations in 1 simulation
        x.dist <- rexp(nosim * nsize, lambda)         # simulated exponential distribution
        x.avg = apply(matrix(x.dist, nosim), 1, mean) # calculates the mean of each simulation
        dat.avg <- data.frame(x = x.avg)              # putting in a data.frame for plotting
        x.avg.mean = mean(dat.avg$x)                  # simulated mean
        x.avg.var = var(dat.avg$x)                    # simulated variance
        x.avg.mean.theoretical = 1/lambda             # theoretical mean
        x.avg.var.theoretical = (1/lambda^2)/nsize    # theoretical variance
```

### Simulated Exponential Distribution With Lambda = 0.2 and 1,000 Observations
```{r echo=FALSE, fig.cap="Figure 1: Exponential Distribution", fig.show='asis'}
createFig1 <- function() {
        
        g <- ggplot(dat, aes(x = x, fill="red")) 
        g <- g + geom_histogram(alpha = .20, binwidth=1, colour = "black") 
        g <- g + geom_vline(xintercept=x.mean) 
        g <- g + annotate(geom = "text", x=10, y=100, label=paste0("simulated mean=", as.character(round(x.mean,digits=2))), size=4, angle=0, vjust=0, hjust=0)
        
        g <- g + annotate(geom = "text", x=10, y=90, label=paste0("theoretical mean=", as.character(round(1/lambda,digits=2))), size=4, angle=0, vjust=0, hjust=0)
        
        g <- g + annotate(geom = "text", x=10, y=130, label=paste0("simulated variance=", as.character(round(x.var, digits=2))), size=4, angle=0, vjust=0, hjust=0)
        
        g <- g + annotate(geom = "text", x=10, y=120, label=paste0("theoretical variance=", as.character(round(1/lambda^2, digits=2))), size=4, angle=0, vjust=0, hjust=0)
        g <- g + theme(legend.position="none")
        g <- g + labs(title = "As expected, the statistics of the simulated distribution are close to the theoretical ")
        print(g)
}
createFig1()
```


### Distribution of Sample Means
``` {r echo=FALSE,  fig.cap="Figure 2: Distribution of Means"}
createFig2 <- function() {
        g <- ggplot(dat.avg, aes(x = x)) 
        
        g <- g + geom_histogram(alpha = .25, binwidth=.2, color = "black", aes(y=..density..)) 
        
        g <- g + geom_vline(aes(xintercept=x.avg.mean, color="Simulated"), size = 1.0) 
        
        g <- g + stat_function(fun=dnorm,args=list(mean=x.avg.mean, sd=sqrt(x.avg.var)), size = 1.0, aes(color="Simulated"))
        
        g <- g + geom_vline(aes(xintercept=x.avg.mean.theoretical, color="Theoretical"), size=.5) 
        
        g <- g + stat_function(fun=dnorm,args=list(mean=x.avg.mean.theoretical, sd=sqrt(x.avg.var.theoretical)), aes(color = "Theoretical"), size = 1.0)
        
        g <- g + labs(title = "Density Distribution of 1,000 Means of 40  Observations", x= " Means of 40 Observations from a Simulated Exponential Distribution")
        print(g)  
}
createFig2()
```

#### Comparison of Simulated and Theoretical Statistics
``` {r echo=FALSE}
statistics <- c("Mean", "Variance")
theoretical <- c(x.avg.mean.theoretical, x.avg.var.theoretical)
simulated <- c(x.avg.mean, x.avg.var)
df <- data.frame(statistics, theoretical, simulated)
names(df) <- c("Statistic", "Theoretical Value", "Simulated Value")
kable(df)
```

### Summary
Comparing Figure 1 to Figure 2 demonstrates how the distribution of means becomes Gaussian for a large number of simulations.  In addition, the distribution of means from the simulation (the green curve and line in Figure 2) is a close approximation to the theoretical normal distribution with mean = 1/lambda and variance = (1/lambda^2) / n, where n is 40 in this simulation (the red curve and line in Figure 2).


# APPENDIX
### Code for Figure 1
``` {r echo=TRUE}
    createFig1
```

### Code for Figure 2
``` {r echo=TRUE}
    createFig2
```
