---
title: 'Statistical Inference: Peer Graded Assigment - Part 2'
author: "Wayne Heller"
date: "April 22, 2017"
output:
  word_document: default
  html_document: default
  pdf_document:
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy.opts = list(width.cutoff=65), tidy=TRUE, fig.pos = 'H', fig.show = 'asis')
```

## Overview
*Part 2 Basic Inferential Data Analysis:*  This part of the analysis uses the ToothGrowth data in the R datasets library and compares tooth growth across dosage and delivery mechanisms.  The specific instructions of this assignemnt are to "Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose"

``` {r echo=FALSE}
        library(ggplot2)
        library(datasets)
```

### Exploratory Plots
``` {r echo=FALSE,  fig.cap="Figure 3: Average Tooth Growth By Dose"}
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
createFig3 <- function() {
        g <- ggplot(data=ToothGrowth, aes(y=len, x=dose, fill=dose)) + geom_boxplot()
        g <- g + labs(title = "Average Tooth Growth Increases with Dose", x="Dose", y="Length")
        print(g)
}
createFig3()

```

``` {r echo=FALSE,  fig.cap="Figure 4: Lower Dose Response"}
createFig4 <- function() {
        g <- ggplot(data=ToothGrowth, aes(y=len, x=supp, fill=supp)) + geom_boxplot()
        g <- g + facet_grid(. ~ dose)
        g <- g + labs(title = "For Lower Doses, Average Growth is Higher for OJ", x="Dose", y="Length")
        print(g)
}
createFig4()

```

### Hypotheses to Test
#### Overall dosage levels, with 95% confidence is Orange Juice a more effective delivery mechanism than Vitamin C
Assumptions: Tooth growth is normally distributed  
Approach: Use Permutation testing to find percentage of results that produce a more extreme difference in averages than was measured  
Ho: There is no significant difference in tooth growth between Orange Juice and Vitamin C
Ha: Average tooth growth is greater for Orange Juice
``` {r echo=TRUE}
y <- ToothGrowth$len
group <- ToothGrowth$supp
testStat <- function(l,g) mean(l[g=="OJ"])- mean(l[g=="VC"])
observedStat <- testStat(y, group)
permutations <- sapply(1:10000, function(i) testStat(y, sample(group)))
observedStat
mean(permutations > observedStat)
```

```{r echo=FALSE,  fig.cap="Figure 5: Permutations Analysis"}
createFig5 <- function() {
        dat <- data.frame(x=permutations)
        g <- ggplot(dat, aes(x = x, fill="red")) 
        g <- g + geom_histogram(alpha = .20, binwidth=1, colour = "black") 
        g <- g + geom_vline(xintercept=observedStat) 
        print(g)
}
createFig5()

```

#### At low dose (0.5), with 95% confidence is Orange Juice a more effective delivery mechanism than Vitamin C
Assumptions: Tooth growth is normally distributed  
Ho: There is no difference in average growth  
Ha: Average tooth growth is greater for Orange Juice

``` {r }
lenOJ <- ToothGrowth$len[ToothGrowth$dose==0.5 & ToothGrowth$supp=="OJ"]
lenVC <- ToothGrowth$len[ToothGrowth$dose==0.5 & ToothGrowth$supp=="VC"]
t.test(lenOJ, lenVC, alternative = "greater", paired = FALSE, var.equal = TRUE)$p.value
```

# APPENDIX
### Code for Figure 3
``` {r echo=TRUE}
    createFig3
```

### Code for Figure 4
``` {r echo=TRUE}
    createFig4
```

### Code for Figure 5
``` {r echo=TRUE}
    createFig5
```